Containers and R: the Rockerverse and beyond

### 2019-11-04 manuscript.Rmd

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    affiliation: Berkeley Institute for Data Science, University of California, Berkeley, CA 94720 USA
    email: karthik.ram@berkeley.edu
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address: "529 Main St, Suite 6610 \\newline Charlestown, MA 02129, USA \\newline ORCiD: 0000-0002-0250-5673"
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
preamble: >
  \usepackage{longtable}
  \usepackage{hyperref}
output: rticles::rjournal_article
#  includes:
#    in_header: preamble.tex
---

@@

_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.

@@

- Binder

The [Binder project](https://mybinder.readthedocs.io/en/latest/), maintained by the team behind Jupyter, makes it possible for users to create and share their computing environments with others. A deployed version of this service called binder hub allows anyone with access to a web browser and an internet connection to launch a temporary instance of these custom environments and execute any notebooks contained within. From a reproducibility standpoint, Binder makes it exceedingly easy to compile a paper, visualize data, and run small examples from papers or tutorials without the need for any local installation.

To set up Binder for a project, a user typically starts at an instance of a Binderhub (examples include mybinder.org and binder.pangeo.io) and passes the location of a hosted Git repository. Using internal tools such as [repo2docker](https://repo2docker.readthedocs.io/en/latest/config_files.html), the binderhub builds a Dockerfile by parsing the contents of all code contained within. While this approach works well for most run of the mill Python projects, it is not so seamless for R projects. For any R projects that use the Tidyverse suite [citation in review], the time and resources required to build all dependencies from source can often time out before completion, making it frustrating for the average R user.

[`holepunch`](https://github.com/karthik/holepunch) is a R package that was designed to remove some of these limitations and make Binder more accessible to novice R users. `holepunch` achieves this by leveraging Rocker images that contains the Tidyverse along special Jupyter dependencies, and only installs additional packages from CRAN and Bioconductor that are not already part of these images. Starting with the Binder/Tidyverse base images eliminates a large part of the build time and in most cases results in a binder instance launching within a minute. `holepunch` as a side effect also creates a DESCRIPTION file which then turns any project into a research compendium [Marwick et al]. The Dockerfile included with the project can also be used to launch a RStudio server locally independent of binder which is especially useful when more computational resources are required.

  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
- GPU [@noamross, @cboettig]
- Gigantum stack
- in education (ready to use teaching/learning environments)
  - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?

>>>>

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820
      - ORCiD \href{https://orcid.org/0000-0001-6419-907X}{0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address: "529 Main St, Suite 6610 \\newline Charlestown, MA 02129, USA \\newline ORCiD: 0000-0002-0250-5673"
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
preamble: >
  \usepackage{longtable}
  \usepackage{hyperref}
output: rticles::rjournal_article
#  includes:
#    in_header: preamble.tex
---

@@

<!--_Do we need a generic intro here?_ DE: Yes, and happy to take a stab.-->

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software.  While there are related approaches such as LXC or Singularity, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.   In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization. By allowing multiple concurrent applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability feature had been realized.  But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.  This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment. [Reference?]   While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too. [Win10; reference?]

The key accomplishment of Docker is to make a "bundled" aggregation of software available to any system running Docker, without requiring much else from the host.  This is a rather attractive, and novel, proposition which has lead to widespread adoption and use of Docker in a variety of domains.  It provided also a natural match for "cloud deployment" which runs (or at least appears to run) "seamlessly" without much explicit reference to the underlying machine, architecture or operating system.  Using containers on cloud hosted servers is a natural fit also because these containers can be deployed with very littler in terms of dependencies on the host system.

For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio Server and Shiny Server. Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN. [Maybe at a later point add forward reference to other topics building on Rocker described here.]

@@

- Binder
  - [`holepunch`](https://github.com/karthik/holepunch) streamlines making an R research compendium Binder-ready [@karthik]
  - [`repo2docker`](https://repo2docker.readthedocs.io/en/latest/config_files.html) (`install.R`, `DESCRPTION`, `runtime.txt`) based on the Rocker images [@nuest]
- GPU [@noamross, @cboettig]
- Gigantum stack
- in education (ready to use teaching/learning environments)
  - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?
